An application of Graph Theory to MDAO problem formulation

### 2012-10-02 intro.tex
starting xdsm for simple

    \begin{align}
        given & \ \ A: {x,y} \rightarrow {z}
        \\      & \ \ B: {x,z} \rightarrow {y^t}
        \\min. &\ \ f(m,t) \notag
        \\w.r.t. & \ \ x,y,z \notag
        \\s.t. & \ \ G(y^t,y) = 0
        \label{eqn:simple}
    \end{align}

    Where $A$ and $B$ represent analysis tools, $F,G$ are the objective and constraint functions respectively. 
    While this does represent a complete problem formulation, 
    it's not the fundamental one since some assumptions have inherently been made about the solution 
    strategy. Analysis $A$ outputs $z$ which is an input to $B$. Hence, $A$ should be run before $B$ with 
    the $y$ being iterated on to convergence. However, a different solution could be equally valid and still represent
    the same fundamental problem:

    \begin{align}
        given & \ \ B: {x,z} \rightarrow {y}
        \\      & \ \ A: {x,y} \rightarrow {z^t}
        \\min. &\ \ f(m,t) \notag
        \\w.r.t. & \ \ x,y,z \notag
        \\s.t. & \ \ G(z^t,z) = 0
        \label{eqn:simple2}
    \end{align}

    Equation \ref{eqn:simple2} differs only slightly from Eqn. \ref{eqn:simple}. $A$ is now dependent on the output of $B$, 
    and $z$ will be iterated on to meet $G$. Now the problem can be solved by running $B$ first and then $A$.
    Since the formulations in Eqn. \ref{eqn:simple} and \ref{eqn:simple_2} both describe the same problem 
    there must be a more fundamental description of the problem that is common between them. We present the FPF as follows:

    \begin{align}
        given & \ \ A: {x,y} \rightarrow {z^t}
        \\      & \ \ B: {x,z} \rightarrow {y^t}
        \\min. &\ \ f(m,t) \notag
        \\w.r.t. & \ \ x,y,z \notag
        \\s.t. & \ \ G1(z^t,z) = 0 \notag
        \\     & \ \ G2(y^t,y) = 0
        \label{eqn:simple_fpf}
    \end{align}

@@

    The challenge with using this traditional mathematical syntax is that it is not easily manipulated or analyzed. 
    A number of matrix based methods have been used successfully to translate the mathematical syntax into a more useful computational form. 
    Steward's Design Structure Matrix (DSM) is a square adjacency matrix which captures the relationship between analysis tools where off 
    diagonal elements of the matrix indicate coupling\cite{Steward1981}. Figure \ref{fig:dsm_simple} shows a DSM for the above notional example. 
    The DSM has been successfully applied in a number of ways to algorithmically modify problem formulations to 
    improve computational efficiency. Rogers et. al developed DeMAID to manipulate a
    DSM  to find an ordering for analysis tools that reduces the cost of solving highly coupled systems\cite{Rogers1996}. This re-ordering yields
    multiple specific problem formulations which all solve the same FPF. In other words, manipulation of the DSM does not fundamentally alter
    the problem formulation, which makes DSM an excellent foundation for specifying the FPF itself. Notice however, that the DSM in 
    Figure \ref{fig:dsm_simple} contains less information than the Eqn. \ref{eqn:simple_fpf}. Objective and constraint information 
    is missing and it's not possible to distinguish individual variables.

    \begin{figure}[!hbp]
        \begin{center}
        \includegraphics[width=.75\textwidth]{images/dsm_simple}
        \caption{Graph and matrix based DSM specifications for Eqn. \ref{eqn:simple_FPF} \label{fig:dsm_simple}}
        \end{center}
    \end{figure}

    Since a DSM describes a square adjacency matrix, it can be represented in an equivalent directed graph where nodes represent analysis tools and 
    edges represent information exchange between those tools. An alternate matrix based syntax, called a 
    Functional Dependence Table (FDT), was proposed by Michelena and Papalambros. 
    FDT represents the relationship between functions, including objectives and constraints, and their values\cite{Michelena1997}. Similar to DSM
    FDT also describes an adjacency matrix of a graph. Unlike the DSM graph, however, a FDT graph is an undirected 
    graph where nodes can represent analysis tools, objectives, or constraints. Edges between nodes represent a dependence on the same 
    variable value. By searching the FDT graph for clusters of totally connected nodes Wagner and Papalambros were identify groups of 
    analysis tools that were all dependent on the same input variables and used that to make partitioning decisions \cite{Wagner1993}. FDT retains 
    a greater portion of the information in the problem formulation, but it is still not complete. It ignores information about feedback coupling 
    between codes.

    It is possible to combine the syntax of an FDT and a DSM by making a small extension to a traditional DSM specification. Normally, a DSM is given 
    with analysis tools as nodes and variable dependencies given as edges. Lamb and Martins included the variables, objectives, and constraint functions
    as nodes in an Extended DSM (XDSM)\cite{Lambe2012} in order to capture a more complete problem formulation for MDAO problems. With XDSM 
    it is possible to specify a complete problem formulation. Lu and Martins developed a process to perform ordering and partitioning on the fundamental 
    problem formulation specified by XDSM in order to achieve significant reductions in computational costs for large scale optimization problems. 
    Figure \ref{fig:dsm_full} illustrates the same notional analysis as in Figure \ref{fig:dsm_simple} represented as an XDSM.

    \begin{figure}[!hbp]
        \begin{center}
        \includegraphics[width=.75\textwidth]{images/dsm_simple}
        \caption{[TODO: Generate XDSM]Extended DSM graph for a notional analysis \label{fig:dsm_full}}
        \end{center}
    \end{figure}

>>>>

    \begin{align}
        given & \ \ A: {x,y} \rightarrow {z}
        \\      & \ \ B: {x,z} \rightarrow {y^t}
        \\min. &\ \ f(m,t) \notag
        \\w.r.t. & \ \ x,y,z \notag
        \\s.t. & \ \ g(y^t,y) = 0
        \label{eqn:simple}
    \end{align}

    Where $A$ and $B$ represent analysis tools, $f,g$ are the objective and constraint functions respectively. 
    While this does represent a complete problem formulation, 
    it's not the fundamental one since some assumptions have inherently been made about the solution 
    strategy. Analysis $A$ outputs $z$ which is an input to $B$. Hence, $A$ should be run before $B$ with 
    the $y$ being iterated on to convergence. However, a different solution could be equally valid and still represent
    the same fundamental problem:

    \begin{align}
        given & \ \ B: {x,z} \rightarrow {y}
        \\      & \ \ A: {x,y} \rightarrow {z^t}
        \\min. &\ \ f(m,t) \notag
        \\w.r.t. & \ \ x,y,z \notag
        \\s.t. & \ \ g(z^t,z) = 0
        \label{eqn:simple2}
    \end{align}

    Equation \ref{eqn:simple2} differs only slightly from Eqn. \ref{eqn:simple}. $A$ is now dependent on the output of $B$, 
    and $z$ will be iterated on to meet $g$. Now the problem can be solved by running $B$ first and then $A$.
    Since the formulations in Eqn. \ref{eqn:simple} and \ref{eqn:simple2} both describe the same problem 
    there must be a more fundamental description of the problem that is common between them. We present the FPF as follows:

    \begin{align}
        given & \ \ A: {x,y} \rightarrow {z^t}
        \\      & \ \ B: {x,z} \rightarrow {y^t}
        \\min. &\ \ f(m,t) \notag
        \\w.r.t. & \ \ x,y,z \notag
        \\s.t. & \ \ g1(z^t,z) = 0 \notag
        \\     & \ \ g2(y^t,y) = 0
        \label{eqn:simple_fpf}
    \end{align}

@@

    The challenge with using this traditional mathematical syntax is that it is not easily manipulated or analyzed. 
    A number of matrix based methods have been used successfully to translate the mathematical syntax into a more useful computational form. 
    Steward's Design Structure Matrix (DSM) is a square adjacency matrix which captures the relationship between analysis tools where off 
    diagonal elements of the matrix indicate coupling\cite{Steward1981}. Figure \ref{fig:dsm_simple} shows a DSM for the above notional example 
    with and without the coupling cycle identified. In order to derive the formulations in Eqn. \ref{eqn:simple} and Eq. \ref{eqn:simple2} 
    you would remove one of the output target variables ($y^t$ or $z^t$) and replace it with a direct connection which would select either $A$
    or $B$ to be executed first. For more complex problems, choosing the proper order to run analysis tools is a non-trival task. 
    Rogers et. al developed DeMAID to manipulate a DSM to find an ordering for analysis tools that 
    reduces the cost of solving highly coupled systems\cite{Rogers1996}. This re-ordering yields multiple specific problem 
    formulations which all solve the same FPF. In other words, manipulation of the DSM does not fundamentally alter
    the problem formulation, which makes DSM an excellent foundation for specifying the FPF itself. Notice however, that the DSM in 
    Figure \ref{fig:dsm_simple} contains less information than the Eqn. \ref{eqn:simple_fpf}. While it does fully capture 
    the coupling information, objective and constraint information is missing.

    \begin{figure}[!hbp]
        \begin{center}
        \includegraphics[height=.25\textheight]{images/dsm_simple}
        \caption{DSM specifications for Eqn. \ref{eqn:simple_fpf}. Coupling cycle is identified by dashed lines. \label{fig:dsm_simple}}
        \end{center}
    \end{figure}

    Since a DSM describes a square adjacency matrix, it can be represented in an equivalent directed graph where nodes represent analysis tools and 
    edges represent information exchange between those tools. An alternate matrix based syntax, called a 
    Functional Dependence Table (FDT), was proposed by Michelena and Papalambros. 
    FDT represents the relationship between functions, including objectives and constraints, and their values\cite{Michelena1997}. Similar to DSM
    FDT also describes an adjacency matrix of a graph. Unlike the DSM graph, however, a FDT graph is an undirected 
    graph where nodes can represent analysis tools, objectives, or constraints. Edges between nodes represent a dependence on the same 
    variable value. By searching the FDT graph for clusters of totally connected nodes Wagner and Papalambros identified groups of 
    analysis tools that were all dependent on the same input variables and used that to make partitioning decisions \cite{Wagner1993}. FDT retains 
    a greater portion of the information in the problem formulation, but it is still not complete. It ignores information about coupling 
    between analyses.

    It is possible to combine the syntax of an FDT and a DSM by making a small extension to a traditional DSM specification. Normally, a DSM is given 
    with analysis tools as nodes and variable dependencies given as edges. Lamb and Martins included the variables, objectives, and constraint functions
    as nodes in an Extended DSM (XDSM)\cite{Lambe2012} in order to capture a more complete problem formulation for MDAO problems. Lu made use of 
    XDSM to perform ordering and partitioning on the fundamental problem formulation specified by XDSM in order to achieve 
    significant reductions in computational costs for large scale optimization problems. Figure \ref{fig:dsm_full} illustrates 
    the same notional analysis as in Figure \ref{fig:dsm_simple} represented as an XDSM.

    \begin{figure}[!hbp]
        \begin{center}
        \includegraphics[width=.75\textwidth]{images/dsm_simple}
        \caption{[TODO: Generate XDSM]Extended DSM graph for Eqn. \ref{eqn:simple_fpf} \label{fig:dsm_full}}
        \end{center}
    \end{figure}
